---
title: "TriLLIEM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TriLLIEM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Initial setup

To get started with using \textbf{TriLLIEM}, we first install and load the package from GitHub via the following commands:

```{r setup}
devtools::install_github("KevinHZhao/TriLLIEM")
library(TriLLIEM)
```

## Simulating triad data

Simulating data in \textbf{TriLLIEM} is all performed through the \texttt{simulateData()} function.
For example, we can simulate $1,000$ case-triads from a population under HWE with a minor allele frequency of $30\%$, $S = 1.2$ and no other genetic factors by using the commands:

```{r simplesim}
set.seed(123) # setting random seed for reproducibility
matdat <- 
  simulateData(
    ntrios = 1000,
    maf = 0.3,
    mtCoef = c(1,1,1),
    S = c(1, 1.2, 1.2^2)
  )$dat4R # extract data formatted for TriLLIEM
## simulateData also returns data formatted for EMIM and Haplin
```

More complex simulations can also be run.
For example, consider the following conditions:

* $1,000$ case-triads
* $1,000$ control-triads
* $30\%$ minor allele frequency
* $(C_1, C_2, C_4) = c(0.5,0.5,0.5)$
* $S=1.2$
* $I_{\textrm{M}} = 1.4$
* $30\%$ environmental exposure frequency in the population
* $V = 1.5$ with maternal imprinting by environment interactions

We can obtain the appropriate simulated data by running:
```{r complexsim}
impdat <-
  simulateData(
    ntrios = 2000,
    maf = 0.3,
    S = c(1, 1.2, 1.2^2),
    V = c(1, 1.5, 1.5^2),
    mtCoef = c(0.5, 0.5, 0.5),
    Im = 1.4,
    includeE = TRUE,
    Einteraction = "Im",
    propE = 0.3,
    includeControl = TRUE,
    nControl = 1000
  )$dat4R
```

## Analyzing triad data

Analyzing triad data sets is performed through the \texttt{TriLLIEM()} function.
This function takes a vector of characters for effects to include in the model, \texttt{effects}.
It also allows users to specify the mating type model to use, and whether to use the stratified or non-stratified approach for environmental interactions (non-stratified by default).
With the first simulated data set above, an appropriate model would use an HWE mating type model with maternal effects only, hence we would run the command:
```{r simpleres}
matres <-
  TriLLIEM(
    mtmodel = "HWE",
    effects = c("M"),
    dat = matdat
  )
```
The \texttt{TriLLIEM()} function returns an object of class \texttt{TriLLIEM}, which inherits from the \texttt{glm} object in R.
This means users can apply many of the usual functions for interpreting results from the \texttt{glm()} function.
To interpret \texttt{matres}, we would run:
```{r simpleinterp}
matres %>% summary() %>% coef()
```
The "Estimate" column for the $\textrm{M}$ row represents our point estimate for $s$, which we can exponentiate to obtain the equivalent risk parameter, $S$.
Based on these results, the model estimates $\hat{S} = 1.24$ with a p-value of $0.00163$.
Since the data were simulated for a population with $S = 1.2$, this level of significance is expected.

Likewise, for the \texttt{impdat} data, we would use an MaS mating type model, with $\textrm{S}, I_{\textrm{M}}, \textrm{E}:\textrm{M}$ as genetic effects in the hybrid model.

```{r complexres}
impres <-
  TriLLIEM(
    mtmodel = "MaS",
    effects = c("M", "Im"),
    dat = impdat,
    includeE = TRUE,
    Einteraction = "Im",
    includeD = TRUE
  )
impres %>% summary() %>% coef()
```
From the results for genetic effects, the model estimates $\hat{\textrm{S}} = 1.19, \hat{I}_{\textrm{M}} = 1.33, \hat{V} = 1.50$ with respective p-values $0.0594, 0.0199, 0.0269$.
Note that these results use the non-stratified approach.
For results equivalent to the stratified approach that are equivalent to \textbf{EMIM} and \textbf{Haplin}, we use:
```{r complexres_strat}
impres_strat <-
  TriLLIEM(
    mtmodel = "MaS",
    effects = c("M", "Im"),
    dat = impdat,
    includeE = TRUE,
    Einteraction = "Im",
    Estrat = TRUE,
    includeD = TRUE
  )
impres_strat %>% summary() %>% coef()
```
This model (which is adds a $\textrm{E}:\text{M}$ effect due to stratification), has $\hat{S} = 1.18$, $\hat{I}_\textrm{M} = 1.34$, and $\hat{V} = 1.46$, with respective p-values $0.0135, 0.0237, 0.0121$.
Note that for comparison with \textbf{EMIM} and \textbf{Haplin}, we can only compare the $\hat{V}$ estimates, since the stratified approach does not produce equivalent estimates for the main effects.
